---
title: "Getting Started with fastMatMR"
author: "Rohit Goswami"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with fastMatMR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

`fastMatMR` is an R package that provides high-performance reading and writing
of Matrix Market files. It wraps around the [fast_matrix_market C++
library](https://github.com/alugowski/fast_matrix_market), ensuring optimal
performance.

## Unique Features

Unlike other packages, such as `Matrix`, `fastMatMR` offers extended support
for:

- **Dense Vectors**: Writing standard R vectors to `.mtx` files.
- **Dense Matrices**: Writing standard R matrices to `.mtx` files.
- **Sparse Matrices**: A more efficient way of reading and writing sparse
  matrices. For performance benchmarks, see performance our vignettes for [write](vignette("fmm_write_bench")) and [read](vignette("fmm_read_bench")) operations.

# Installation

To install the development version of `fastMatMR` from GitHub:

```{r, eval=FALSE}
# install.packages("devtools")
devtools::install_github("HaoZeke/fastMatMR")
```

# Basic Usage

Load the `fastMatMR` package:

```{r}
library(fastMatMR)
```

## Writing Matrix Market files

### With Dense Vectors (Not Supported by `Matrix`)

```{r}
vec <- c(1, 2, 3)
write_fmm(vec, "vector.mtx")
```

### With Dense Matrices (Not Supported by `Matrix`)

```{r}
mat <- matrix(c(1, 2, 3, 4), nrow = 2)
write_fmm(mat, "matrix.mtx")
```

### With Sparse Matrices

```{r}
sp_mat <- Matrix::sparseMatrix(i = c(1, 3), j = c(2, 4), x = 7:8)
write_fmm(sp_mat, "sparse.mtx")
```

## Reading Matrix Market files

- `fastMatMR` will correctly roundtrip `NaN` values.
- `NA` values are not supported by the Matrix Market format, and are coerced to `NaN`

### With Dense Vectors

This is not supported by `Matrix`.

```{r}
vec <- c(1, 2, 3.32, NaN)
vec_to_fmm(vec, "vector.mtx")
fmm_to_vec("vector.mtx")
```


### Addenum

Consider:

```{r}
Matrix::Matrix(c(1, 0, 3, NA), nrow = 2, sparse = TRUE) ->spmat
Matrix::writeMM(spmat, "sparse_na.mtx")
Matrix::readMM("sparse_na.mtx")
```

Which returns:

```{r eval=FALSE}
2 x 2 sparse Matrix of class "dgTMatrix"
             
[1,] 1  3e+00
[2,] . 1e+308
```

## Reading Back In Python

You can read the `.mtx` files back in Python:

```bash
pip install fast_matrix_market
python -c 'import fast_matrix_market as fmm; print(fmm.read_array_or_coo("sparse.mtx"))'
```


# Summary

`fastMatMR` offers unique capabilities in the R ecosystem for working with
Matrix Market files, including support for dense vectors and matrices. It also
provides high-performance sparse matrix operations, as benchmarked in our
performance vignettes for [write](vignette("fmm_write_bench")) and [read](vignette("fmm_read_bench")) operations.
